---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r do_not_add_numbers_in_image_filenames, include=FALSE}
knitr::opts_chunk$set(fig.process = function(x) {
  x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
  if (file.rename(x, x2)) x2 else x
})
```

## Loading and preprocessing the data
* load the data
* transform the data into a format suitable for the analysis

```{r}
unzip("activity.zip")
df <- read.csv("activity.csv")
df <- df[complete.cases(df), ]
summary(df)
```

## What is mean total number of steps taken per day?

* Calculate the total number of steps taken per day

```{r}
steps_per_day <- aggregate(steps ~ date, df, sum)
steps_per_day
```

* Make a histogram of the total number of steps taken each day

```{r 1_steps_per_day}
hist(
    steps_per_day$steps,
    breaks=5,
    xlab="Number of steps taken in a day", 
    main="Histogram of the total number of steps taken each day")
```

* Calculate and report the mean and median of the total number of steps
  taken per day

```{r}
m1 <- round(mean(steps_per_day$steps), 2)
m2 <- round(median(steps_per_day$steps), 2)
print(paste("The mean and median number of steps taken per day:", m1, m2))
```

## What is the average daily activity pattern?

* Make a time series plot of the 5-minute interval (x-axis) and the average
  number of steps taken, averaged across all days (y-axis)

```{r 2_steps_per_interval_before_imputting_missing_values}
mean_activity <- aggregate(steps ~ interval, df, mean)

plot(
    mean_activity$interval, 
    mean_activity$steps, 
    type="l",
    xlab="Time interval",
    ylab="Average number of steps",
    main="The average number of steps taken\nin each 5-minute interval")
```

* Which 5-minute interval, on average across all the days in the dataset, 
contains the maximum number of steps?

```{r}
max_activity_interval <- mean_activity[which.max(mean_activity$steps),]
print(paste(
    "The interval with the maximum average number of steps across all days:",
    "interval =", max_activity_interval$interval,
    "steps =", round(max_activity_interval$steps, 0)))
```

## Imputing missing values

* Calculate and report the total number of missing values in the dataset
    * the only variable containing missing values is *steps*

```{r}
ndf <- read.csv('activity.csv', header=TRUE)
print(paste(
    "The number of entries with missing values is:", sum(is.na(ndf$steps))))
```

* Create a new dataset that is equal to the original dataset but with the 
  missing data filled in
    * devise a strategy for filling in all of the missing values in the dataset
    * the strategy does not need to be sophisticated
    * for example, you could use the mean/median for that day, 
      or the mean for that 5-minute interval, etc


```{r}
activity <- aggregate(steps ~ interval, ndf, mean)

nchanges <- 0
for (i in 1:nrow(ndf)){
    if (is.na(ndf$steps[i])){
        cinterval <- ndf$interval[i]
        ndf$steps[i] <- round(activity$steps[activity$interval==cinterval], 2)
        nchanges <- nchanges + 1
    }
}
print(paste("Updated", nchanges, "missing values in the 'steps' variable"))
```

* Make a histogram of the total number of steps taken each day 

```{r 3_steps_per_interval_after_imputting_missing_values}
n_steps_per_day <- aggregate(steps ~ date, ndf, sum)

hist(
    n_steps_per_day$steps,
    breaks=5,
    xlab="Number of steps taken in a day", 
    main="Histogram of the total number of steps taken each day")
```

* Calculate and report the mean and median total number of steps taken per day.

```{r}
nm1 <- round(mean(n_steps_per_day$steps), 2)
nm2 <- round(median(n_steps_per_day$steps), 2)
print(paste("The mean and median number of steps taken per day:", nm1, nm2))
```

* Do these values differ from the estimates from the 1st part of the assignment? 

```{r}
d1 <- round(m1 - nm1, 2)
d2 <- round(m2 - nm2, 2)
print(paste(
    "Difference in mean and median estimates after imputting missing values", 
    d1, d2))
```

* What is the impact of imputing missing data on the estimates 
  of the total daily number of steps?

```{r}
df <- read.csv('activity.csv', header=TRUE)
steps_per_day <- aggregate(steps ~ date, df, sum, na.action=na.pass)

differences <- data.frame(
    referenceSteps=steps_per_day$steps, 
    updatedSteps=n_steps_per_day$steps,
    differenceSteps=n_steps_per_day$steps - steps_per_day$steps)

print(differences)
```

## Are there differences in activity patterns between weekdays and weekends?

* Create a new factor variable in the dataset with two levels -
  "weekday" and "weekend" indicating whether a given date
  is a weekday or weekend day

```{r}
get_day_type <- function(given_date){
    day_name <- weekdays(as.Date(given_date))
    day_type <- NA
    if (day_name %in% c("Saturday", "Sunday")){
        day_type <- "weekend"
    }
    else{
        day_type <- "weekday"
    }
    day_type
}

ndf$daytype <- as.factor(sapply(ndf$date, get_day_type))

print(table(weekdays(as.Date(ndf$date))))
print(summary(ndf$daytype))
```

* Make a panel plot containing a time series plot of the 5-minute interval 
  (x-axis) and the average number of steps taken, 
  averaged across all weekday days or weekend days (y-axis).

```{r 4_steps_per_interval_by_day_type, warning=FALSE}
library(ggplot2)
average_steps_per_interval <- aggregate(steps ~ interval + daytype, ndf, mean)

ggplot(data=average_steps_per_interval, aes(x=interval, y=steps)) + 
    geom_line(aes(color=daytype)) +
    labs(x="Interval", y="Average number of steps") +
    facet_wrap( ~ daytype, nrow = 2) +
    theme_classic() +
    theme(
        legend.position = "none", 
        strip.background=element_rect(fill="bisque"))
```
