---
output: 
  html_document:
    keep_md: true
---

# Health and economic impact of weather events

## Introduction

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities.
Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database.
This database tracks characteristics of major storms and weather events in the United States,
including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

## Synopsis

The goal of this analysis is to identify the health and economic impact of weather events.

Two data processing steps are necessary for this analysis:
* cleaning the event names
* reconstructing the complete values of property damage and crop damage.

The analysis reveals that the most dangerous event for the population health are the tornados, causing 37% of the fatalities.
They are followed by the event of excessive heat which is causing 13% of fatalities.

The analysis also reveals that the event causing the greatest economic consequences are the flash floods, generating 75% of the damage costs.
They are followed by the event of thunderstorm winds which is generating 23% of expenses.

The most severe weather event with respect to crop damage is the drought.
It generates 28% of the damage costs.

## Details of the analysis

### Data description

The data is presented under the format of a csv file compressed via the bzip2 algorithm.
The achive can be downloaded from the [course](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) web site.

The documentation is available at:
* National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
* National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

The events in the database start in the year 1950 and end in November 2011.
In the earlier years of the database there are generally fewer events recorded,
most likely due to a lack of good records.
More recent years should be considered more complete.

### Goal of analysis

The goal is to explore the NOAA Storm Database and answer some basic questions about severe weather events:
* across the United States, which types of events are most harmful with respect to population health?
* across the United States, which types of events have the greatest economic consequences?

### Data processing

* download the archive and read the data

```{r}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
archive <- "StormData.csv.bz2"

if ( !file.exists(archive)){
    download.file(url, archive)
}

data <- read.csv(bzfile(archive), header=TRUE)
```

* inspect the data

```{r}
str(data)
```

* select only the variables that are useful for the analysis
    * event type
    * population health: fatalities and injuries
    * economic consequences: property damage and crop damage
    
```{r}
data <- data[, c(8, 23, 24, 25, 26, 27, 28)]
str(data)
```

* clean the event names
    * upercase the name of events
    * replace all punctuation marks with spaces
    * trim leading and trailing whitespaces

```{r}
events <- toupper(data$EVTYPE)
events <- gsub("[[:punct:]]", " ", events)
events <- gsub("^[[:space:]]+|[[:space:]]+$", "", events)
print(paste(
    "Passed from", length(unique(data$EVTYPE)), 
    "to", length(unique(events)), "unique event names"))
data$EVTYPE <- events
```

* reconstruct the property damage and crop damage values
    * combine *PROPDMG* (floating value) and *PROPDMGEXP* (exponent) variables
    * combine *CROPDMG* (floating value) and *CROPDMGEXP* (exponent) variables

```{r}
compose_value <- function(exp){
    if (exp %in% c('h', 'H'))
        return(10 ^ 2)
    else if (exp %in% c('k', 'K'))
        return(10 ^ 3)
    else if (exp %in% c('m', 'M'))
        return(10 ^ 6)
    else if (exp %in% c('b', 'B'))
        return(10 ^ 9)
    else if (!is.na(as.numeric(exp)))
        return(10 ^ as.numeric(exp))
    else if (exp %in% c('', '-', '?', '+'))
        return(1)
    NA
}

data$PROPDMG <- data$PROPDMG * sapply(data$PROPDMGEXP, compose_value)
data$CROPDMG <- data$CROPDMG * sapply(data$CROPDMGEXP, compose_value)
data <- data[, c(1, 2, 3, 4, 6)]

str(data)
```

### Explore the consequences of severe weather events

* identify which types of events are most harmful with respect to population health?

    * inspect the injuries counts with respect to the event types

    ```{r}
    hdata1 <- aggregate(INJURIES ~ EVTYPE, data, sum)
    hdata1 <- hdata1[order(-hdata1$INJURIES),]
    print("The top 10 events that caused the most number of injuries:")
    head(hdata1, 10)
    ```

    * inspect the fatalities counts with respect to the event types

    ```{r}
    hdata2 <- aggregate(FATALITIES ~ EVTYPE, data, sum)
    hdata2 <- hdata2[order(-hdata2$FATALITIES),]
    print("The top 10 events that caused the most number of fatalities:")
    head(hdata2, 10)
    ```

    * inspect percentages of most dangerous events

    ```{r}
    total_fatalities <- sum(hdata2$FATALITIES)

    for (event in head(hdata2$EVTYPE, 3)){
        n <- sum(hdata2[which(hdata2$EVTYPE == event), ]$FATALITIES)
        p <- round(n / total_fatalities * 100, 2)
        print(paste("The", event, "causes", p, "% of fatalities"))
    }    
    ```

    * inspect the injuries and fatalities counts with a plot 

    ```{r}
    hdata1 <- head(hdata1, 10)
    hdata2 <- head(hdata2, 10)
    ```

    ```{r population_health, fig.width=18, fig.height=6}
    par(mfrow=c(1:2))
    par(mar=c(4, 13, 2, 2))
    options(scipen = 50) 

    barplot(
        hdata1$INJURIES, 
        names.arg=hdata1$EVTYPE, 
        horiz=TRUE,
        las=1,
        col="blue",
        xlim=c(0, 100000),
        xlab="Number of reported injuries",
        ylab="Event Type\n\n\n\n\n\n\n\n",
        main="Top 10 causes of injuries") 
    barplot(
        hdata2$FATALITIES, 
        names.arg=hdata2$EVTYPE, 
        horiz=TRUE,
        las=1,
        col="red", 
        xlim=c(0, 6000),
        xlab="Number of reported fatalities",
        ylab="Event Type\n\n\n\n\n\n",
        main="Top 10 causes of fatalities") 
    ```


* identify which types of events have the greatest economic consequences?

    * inspect the property damage with respect to the event types

    ```{r}
    ecdata1 <- aggregate(PROPDMG ~ EVTYPE, data, sum)
    ecdata1 <- ecdata1[order(-ecdata1$PROPDMG),]
    print("The top 10 events that caused the greatest property damage:")
    head(ecdata1, 10)
    ```

    * inspect the crop damage with respect to the event types

    ```{r}
    ecdata2 <- aggregate(CROPDMG ~ EVTYPE, data, sum)
    ecdata2 <- ecdata2[order(-ecdata2$CROPDMG),]
    print("The top 10 events that caused the greatest crop damage:")
    head(ecdata2, 10)
    ```

    * inspect percentages of most property damaging events

    ```{r}
    total_expenses <- sum(ecdata1$PROPDMG)

    for (event in head(ecdata1$EVTYPE, 3)){
        n <- sum(ecdata1[which(ecdata1$EVTYPE == event), ]$PROPDMG)
        p <- round(n / total_expenses * 100, 2)
        print(paste("The", event, "causes", p, "% of expenses"))
    }    
    ```
    
    
    * inspect percentages of most crop damaging events

    ```{r}
    total_expenses <- sum(ecdata2$CROPDMG)

    for (event in head(ecdata2$EVTYPE, 3)){
        n <- sum(ecdata2[which(ecdata2$EVTYPE == event), ]$CROPDMG)
        p <- round(n / total_expenses * 100, 2)
        print(paste("The", event, "causes", p, "% of expenses"))
    }    
    ```

    * inspect the economic consequences with a plot 

    ```{r}
    ecdata1 <- head(ecdata1, 10)
    ecdata2 <- head(ecdata2, 10)
    ecdata1$PROPDMG <- ecdata1$PROPDMG / (10^9)
    ecdata2$CROPDMG <- ecdata2$CROPDMG / (10^9)
    ```

    ```{r economic_consequences, fig.width=18, fig.height=6}
    par(mfrow=c(1:2))
    par(mar=c(4, 13, 2, 2))
    options(scipen = 3) 

    barplot(
        ecdata1$PROPDMG, 
        names.arg=ecdata1$EVTYPE, 
        horiz=TRUE,
        las=1,
        col="blue",
        xlim=c(0, 70000),
        xlab="Expenses generated [billions of dollars]",
        ylab="Event Type\n\n\n\n\n\n\n\n\n",
        main="Top 10 causes of property damage")
    
    barplot(
        ecdata2$CROPDMG, 
        names.arg=ecdata2$EVTYPE, 
        horiz=TRUE,
        las=1,
        col="red",
        xlim=c(0, 14),
        xlab="Expenses generated [billions of dollars]",
        ylab="Event Type\n\n\n\n\n\n\n\n\n",
        main="Top 10 causes of crop damage")
    ```

### Results

The greatest number of deaths and injuries are caused by tornados,
amounting to more than 5,000 deaths and to more than 90,000 injuries.
Other dangerous event types include the excessive heat, floods and winds.

The most property damages are caused by the flash floods and are estimated at 68 trillion dollars.
The thunderstorm winds come in second place causing 20 trillion dollars worth of damages.

The most severe weather event in terms of crop damage is the drought. 
It has caused more than 13 billion dollars worth of damages.
Other damaging events include the floods and the hails.
